---
title: "Themes"
author: "Justin Baumann"
format: 
  html:
    toc: true
  pdf:
    toc: true
    number-sections: true
    colorlinks: true
    self-contained: true

editor: visual
---

```{r}
#| warning: false
#Load packages
library(tidyverse)
library(ggsci) #for easy color scales
library(patchwork) #to make multi-panel plots 
library(palmerpenguins) # our fave penguin friends :)
```

## **Themes**

Themes allow us to change the background color and most other aspects of a plot. There are a range of theme options within ggplot that will allow us to quickly make clean plots. The two that are most commonly used are theme_bw() and theme_classic()

**Default theme** (with terrible gray background)

```{r}
ggplot(data=penguins, aes(x=species, y= bill_length_mm)) +
  geom_boxplot(aes(fill=species))+
  scale_fill_aaas()+
  labs(x = 'Species', y='Bill length (mm)', title='Penguin bill length by species')
```

**theme_bw()** (removes gray background)

```{r}
ggplot(data=penguins, aes(x=species, y= bill_length_mm)) +
  geom_boxplot(aes(fill=species))+
  scale_fill_aaas()+
  labs(x = 'Species', y='Bill length (mm)', title='Penguin bill length by species')+
  theme_bw()
```

**theme_classic()** (removes gray and grid lines)

```{r}
ggplot(data=penguins, aes(x=species, y= bill_length_mm)) +
  geom_boxplot(aes(fill=species))+
  scale_fill_aaas()+
  labs(x = 'Species', y='Bill length (mm)', title='Penguin bill length by species')+
  theme_classic()
```

The theme() function in ggplot is SUPER flexible. You can pretty much do anything with it. This is key for customizing plots. I'd encourage you to play around with this a bit. [Here](https://ggplot2.tidyverse.org/reference/theme.html) is a great place to learn more and see examples.\
\##**Some examples of using theme()**

**Changing text size**

```{r}
ggplot(data=penguins, aes(x=species, y= bill_length_mm)) +
  geom_boxplot(aes(fill=species))+
  scale_fill_aaas()+
  labs(x = 'Species', y='Bill length (mm)', title='Penguin bill length by species')+
  theme(text=element_text(size=24))
```

**Remove the gray background**

```{r}
ggplot(data=penguins, aes(x=species, y= bill_length_mm)) +
  geom_boxplot(aes(fill=species))+
  scale_fill_aaas()+
  labs(x = 'Species', y='Bill length (mm)', title='Penguin bill length by species')+
  theme(text=element_text(size=24), panel.background = element_rect(fill="white")) #can use any color
```

**Turn the X-Axis text**

```{r}
ggplot(data=penguins, aes(x=species, y= bill_length_mm)) +
  geom_boxplot(aes(fill=species))+
  scale_fill_aaas()+
  labs(x = 'Species', y='Bill length (mm)', title='Penguin bill length by species')+
  theme(text=element_text(size=24), panel.background = element_rect(fill="white"), axis.text.x=element_text(angle=90, vjust=0.5, hjust=0.8)) #can adjust vertical and horizontal text positions
```
